---
title: "Vignette"
author: "Shiyao Xu"
date: "1/19/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
devtools::install_github("zeger-nishimura-lab/CausalSurvival", auth_token = "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")
```

```{r}
library(tidyverse)
library(glmnet)
library(CausalSurvival)
```

```{r}
hazMethod <- "ns"
```

## Load data

Use synthetic `Eunomia` dataset to illustrate how to perform the simulation study. 

```{r message=FALSE}

connectionDetails <- Eunomia::getEunomiaConnectionDetails()
Eunomia::createCohorts(connectionDetails)

cohortMethodData <- CohortMethod::getDbCohortMethodData(connectionDetails = connectionDetails,
                                                        cdmDatabaseSchema = "main",
                                                        targetId = 1,
                                                        comparatorId = 2,
                                                        outcomeIds = 3,
                                                        exposureDatabaseSchema = "main",
                                                        outcomeDatabaseSchema = "main",
                                                        exposureTable = "cohort",
                                                        outcomeTable = "cohort",
                                                        covariateSettings = FeatureExtraction::createDefaultCovariateSettings())

population <- CohortMethod::createStudyPopulation(cohortMethodData = cohortMethodData,
                                                  outcomeId = 3,
                                                  riskWindowEnd = 99999)

covariateData <- CausalSurvival::filterAndTidyCovariatesForPs(cohortMethodData, population)
covariates <- covariateData$covariates
row_id <- covariates %>% dplyr::pull(.data$rowId)
covariate_id <- covariates %>% dplyr::pull(.data$covariateId)
val <- covariates %>% dplyr::pull(.data$covariateValue)
covariates <- data.frame(rowId=row_id, covariateId=covariate_id, covariateValue=val)
covariates <- covariates[-which(covariates$covariateId %in% c(1118084410, 1118084412, 1118084413, 1124300410, 1124300412, 1124300413)), ]
covariates$covariateId <- as.numeric(factor(covariates$covariateId))

treatment <- population$treatment
event_time <- population$survivalTime
id <- 1:length(treatment)
rm(list=c("row_id", "covariate_id", "val", "covariateData", "cohortMethodData", "connectionDetails", "population"))
```

Simulate censoring time separately for each treatment group. 

```{r}
censoring_time <- 6 + simCenTime(treatment=treatment, covariates=covariates, 
                                 lambda1=0.000003, nu1=1.5, lambda0=0.000003, nu0=1.5, seed=2019)
outcome <- 1*(event_time <= censoring_time)
time <- apply(cbind(event_time, censoring_time), 1, min)
```



## Compute on real data

### PS calculation from real data

```{r}
treatProb <- CausalSurvival::estimateTreatProb(id=id, treatment=treatment, covariates=covariates)
```

### Survival and censoring hazards from real data

First perform pre-screening of the baseline covariates via cox LASSO for both survival and censoring hazards. 

```{r}
cov <- Matrix::sparseMatrix(i = covariates$rowId, j = covariates$covariateId, x = covariates$covariateValue, repr = "T")

set.seed(1)
survFit <- glmnet::cv.glmnet(x=cbind(treatment, cov), y=survival::Surv(time=time, event=outcome), 
                             family = "cox", nfolds = 5, penalty.factor = c(0, rep(1, dim(cov)[2])))
survCf <- coef(survFit, s = survFit$lambda.1se)
survCov_indx <- setdiff(which(survCf != 0)-1, 0)

set.seed(0)
cenFit <- glmnet::cv.glmnet(x=cbind(treatment, cov), y=survival::Surv(time=time, event=1-outcome), 
                            family = "cox", nfolds = 5, penalty.factor = c(0, rep(1, dim(cov)[2])))
cenCf <- coef(cenFit, s = cenFit$lambda.1se)
cenCov_indx <- setdiff(which(cenCf != 0)-1, 0)

```

Then perform ridge regression with the selected baseline covariates, treatment, functions of time, interaction between treatment and functions of time to predict survival and censoring hazards. Time is coarsened to non-uniform time intervals with each interval containing about the same number of events. 

```{r}

coarsenedData <- CausalSurvival::coarsenData(time=time, outcome=outcome)

estimatedSurvHazard <- CausalSurvival::estimateHazards(coarsenedData=coarsenedData, outcome=outcome, 
                                                       treatment=treatment, covariates=covariates, 
                                                       covId=survCov_indx,
                                                       hazEstimate="survival", hazMethod=hazMethod)

estimatedCenHazard <- CausalSurvival::estimateHazards(coarsenedData=coarsenedData, outcome=outcome, 
                                                      treatment=treatment, covariates=covariates, 
                                                      covId=cenCov_indx,
                                                      hazEstimate="censoring", hazMethod=hazMethod)

```


### Algorithm 

Before conducting the simulation studies, let's first see how to run the algorithm with the above estimated survival and censoring hazards. 

```{r}
tmle <- CausalSurvival::computeEstimator(treatment=treatment, outcome=outcome, time=coarsenedData$timeInt,
                                         initial_survHaz=estimatedSurvHazard$haz, cenHaz=estimatedCenHazard$haz, treatProb=treatProb,
                                         coarseningParam=coarsenedData, method="TMLE")

aipw <- CausalSurvival::computeEstimator(treatment=treatment, outcome=outcome, time=coarsenedData$timeInt,
                                         initial_survHaz=estimatedSurvHazard$haz, cenHaz=estimatedCenHazard$haz, treatProb=treatProb,
                                         coarseningParam=coarsenedData, method="AIPW")

ipw <- CausalSurvival::computeEstimator(treatment=treatment, outcome=outcome, time=coarsenedData$timeInt,
                                        initial_survHaz=estimatedSurvHazard$haz, cenHaz=estimatedCenHazard$haz, treatProb=treatProb,
                                        coarseningParam=coarsenedData, method="IPW")

cox <- CausalSurvival::computeEstimator(treatment=treatment, outcome=outcome, time=coarsenedData$timeInt,
                                        initial_survHaz=estimatedSurvHazard$haz, cenHaz=estimatedCenHazard$haz, 
                                        treatProb=treatProb, coarseningParam=coarsenedData_sim, 
                                        method="cox", hazMethod=hazMethod)

weightedCox <- CausalSurvival::computeEstimator(treatment=treatment, outcome=outcome, time=coarsenedData$timeInt,
                                                initial_survHaz=estimatedSurvHazard$haz, cenHaz=estimatedCenHazard$haz, 
                                                treatProb=treatProb, coarseningParam=coarsenedData_sim, 
                                                method="weightedCox", hazMethod=hazMethod)

```


### Result

#### Survival curve and risk difference

```{r}
par(mfrow=c(1,2))
plot(coarsenedData$breaks, c(1, ipw$S1), main="S(1)", xlab="Time (90 days)", ylab="Survival probability", type="l", ylim=c(0.3, 1), lwd=1, col=367)
lines(coarsenedData$breaks, c(1, aipw$S1), col=562)
lines(coarsenedData$breaks, c(1, tmle$S1), col=413)
lines(coarsenedData$breaks, c(1, cox$S1), col=547)
lines(coarsenedData$breaks, c(1, weightedCox$S1), col="green4")
plot(coarsenedData$breaks, c(1, ipw$S0), col=367, main="S(0)", xlab="Time (90 days)", ylab="Survival probability", type="l", ylim=c(0.3, 1), lwd=1)
lines(coarsenedData$breaks, c(1, aipw$S0), col=562)
lines(coarsenedData$breaks, c(1, tmle$S0), col=413)
lines(coarsenedData$breaks, c(1, cox$S0), col=547)
lines(coarsenedData$breaks, c(1, weightedCox$S0), col="green4")
```


```{r}
par(mfrow=c(1,2))
plot(coarsenedData$breaks, c(0, ipw$risk_diff), main="Risk difference", xlab="Time (days)", ylab="Difference in survival", type = "b", pch = 20, ylim=c(-0.2, 0.1), lwd=1, col=367)
lines(coarsenedData$breaks, c(0, aipw$risk_diff), col=562, type = "b", pch = 20)
lines(coarsenedData$breaks, c(0, tmle$risk_diff), col=413, type = "b", pch = 20)
lines(coarsenedData$breaks, c(0, cox$risk_diff), col=547, type = "b", pch = 20)
lines(coarsenedData$breaks, c(0, weightedCox$risk_diff), col="green4", type = "b", pch = 20)
legend("bottomleft", c("IPW", "Augmented IPW", "TMLE", "Stratified cox", "Weighted stratified cox"),col=c(367, 562, 413, 547, "green4"), box.lty=0, lwd=c(1, 1, 1, 1, 1), cex=0.6)
plot(coarsenedData$breaks[-1], ipw$SE_risk_diff, main="SD of risk difference", xlab="Time (days)", ylab="Standard deviation", type = "p", pch = 20, ylim=c(0, 0.09), col=367)
lines(coarsenedData$breaks[-1], aipw$SE_risk_diff, col=562, type = "p", pch = 20)
lines(coarsenedData$breaks[-1], tmle$SE_risk_diff, col=413, type = "p", pch = 20)
lines(coarsenedData$breaks[-1], cox$SE_risk_diff, col=547, type = "p", pch = 20)
lines(coarsenedData$breaks[-1], weightedCox$SE_risk_diff, col="green4", type = "p", pch = 20)
```

#### Difference in RMST

```{r}
par(mfrow=c(1,2))
plot(coarsenedData$breaks, c(0, ipw$rmst_diff), main="Difference in RMST", xlab="Time (days)", ylab="Difference in RMST", type = "b", pch = 20, ylim=c(-0.2, 0.1), lwd=1, col=367)
lines(coarsenedData$breaks, c(0, aipw$rmst_diff), col=562, type = "b", pch = 20)
lines(coarsenedData$breaks, c(0, tmle$rmst_diff), col=413, type = "b", pch = 20)
lines(coarsenedData$breaks, c(0, cox$rmst_diff), col=547, type = "b", pch = 20)
lines(coarsenedData$breaks, c(0, weightedCox$rmst_diff), col="green4", type = "b", pch = 20)
legend("bottomleft", c("IPW", "Augmented IPW", "TMLE", "Stratified cox", "Weighted stratified cox"),col=c(367, 562, 413, 547, "green4"), box.lty=0, lwd=c(1, 1, 1, 1, 1), cex=0.6)
plot(coarsenedData$breaks[-1], ipw$SE_rmst_diff, main="SD of difference in RMST", xlab="Time (days)", ylab="Standard deviation", type = "p", pch = 20, ylim=c(0, 0.09), col=367)
lines(coarsenedData$breaks[-1], aipw$SE_rmst_diff, col=562, type = "p", pch = 20)
lines(coarsenedData$breaks[-1], tmle$SE_rmst_diff, col=413, type = "p", pch = 20)
lines(coarsenedData$breaks[-1], cox$SE_rmst_diff, col=547, type = "p", pch = 20)
lines(coarsenedData$breaks[-1], weightedCox$SE_rmst_diff, col="green4", type = "p", pch = 20)
```

## Simulation studies

### Simulation 

Simulate one set of data with the estimated survival and censoring hazard. 

```{r}
simData <- CausalSurvival::simulateData(treatment=treatment, survHaz=estimatedSurvHazard$haz, cenHaz=estimatedCenHazard$haz,
                                        coarsenedTime=coarsenedData, seed=2019)
```

### Survival and censoring hazards from simulated data

Then perform ridge regression with the selected baseline covariates, treatment, functions of time, interaction between treatment and functions of time to predict survival and censoring hazards in the simulated data. 

```{r}
## replace timeInt with simulated data
coarsenedData_sim <- coarsenedData
coarsenedData_sim$timeInt <- simData$ObservedTime

## survival hazards and selected covariates
estimatedSurvHazard_sim <- CausalSurvival::estimateHazards(coarsenedData=coarsenedData_sim, outcome=simData$ObservedEvent, 
                                                           treatment=treatment, covariates=covariates, 
                                                           covId=survCov_indx,
                                                           hazEstimate="survival", hazMethod=hazMethod)
## censoring hazards and selected covariates
estimatedCenHazard_sim <- CausalSurvival::estimateHazards(coarsenedData=coarsenedData_sim, outcome=simData$ObservedEvent, 
                                                          treatment=treatment, covariates=covariates, 
                                                          covId=cenCov_indx,
                                                          hazEstimate="censoring", hazMethod=hazMethod)

```


### Algorithm 

Run the algorithm with the simulated data and estimated hazards. 

```{r}
tmle_sim <- CausalSurvival::computeEstimator(treatment=treatment, outcome=simData$ObservedEvent, time=simData$ObservedTime,
                                             initial_survHaz=estimatedSurvHazard_sim$haz, cenHaz=estimatedCenHazard_sim$haz, treatProb=treatProb,
                         coarseningParam=coarsenedData, method="TMLE")

aipw_sim <- CausalSurvival::computeEstimator(treatment=treatment, outcome=simData$ObservedEvent, time=simData$ObservedTime,
                                             initial_survHaz=estimatedSurvHazard_sim$haz, cenHaz=estimatedCenHazard_sim$haz, treatProb=treatProb,
                                             coarseningParam=coarsenedData, method="AIPW")

ipw_sim <- CausalSurvival::computeEstimator(treatment=treatment, outcome=simData$ObservedEvent, time=simData$ObservedTime,
                                            initial_survHaz=estimatedSurvHazard_sim$haz, cenHaz=estimatedCenHazard_sim$haz, treatProb=treatProb,
                                            coarseningParam=coarsenedData, method="IPW")

cox_sim <- CausalSurvival::computeEstimator(treatment=treatment, outcome=simData$ObservedEvent, time=simData$ObservedTime,
                                            initial_survHaz=estimatedSurvHazard_sim$haz, cenHaz=estimatedCenHazard_sim$haz, 
                                            treatProb=treatProb, coarseningParam=coarsenedData_sim, 
                                            method="cox", hazMethod=hazMethod)

weightedCox_sim <- CausalSurvival::computeEstimator(treatment=treatment, outcome=simData$ObservedEvent, time=simData$ObservedTime,
                                                    initial_survHaz=estimatedSurvHazard_sim$haz, cenHaz=estimatedCenHazard_sim$haz, 
                                                    treatProb=treatProb, coarseningParam=coarsenedData_sim, 
                                                    method="weightedCox", hazMethod=hazMethod)

```


### Counterfactual survival curve and RMST

Derive counterfactual survival curve from the estimated survival hazard. And transform to RMST. 

```{r}
Truth_S <- CausalSurvival::calculateSurvCurve(coarsenedTime=coarsenedData, survHaz=estimatedSurvHazard$haz)
Truth_rmst <- CausalSurvival::calculateRMST(coarsenedTime=coarsenedData, survCurve=Truth_S)
```

### Result

#### Survival curves and risk difference

```{r}
par(mfrow=c(1,2))
plot(coarsenedData$breaks, c(1, Truth_S$S1), main="S(1)", xlab="Time (90 days)", ylab="Survival probability", type="l", ylim=c(0.3, 1), lwd=2, col=24)
lines(coarsenedData$breaks, c(1, ipw_sim$S1), col=367)
lines(coarsenedData$breaks, c(1, aipw_sim$S1), col=562)
lines(coarsenedData$breaks, c(1, tmle_sim$S1), col=413)
lines(coarsenedData$breaks, c(1, cox_sim$S1), col=547)
lines(coarsenedData$breaks, c(1, weightedCox_sim$S1), col="green4")
legend("bottomleft", c("Truth","IPW", "Augmented IPW", "TMLE", "Stratified cox", "Weighted stratified cox"),col=c(24, 367, 562, 413, 547, "green4"), box.lty=0, lwd=c(2, 1, 1, 1, 1, 1), cex=0.6)
plot(coarsenedData$breaks, c(1, Truth_S$S0), main="S(0)", xlab="Time (90 days)", ylab="Survival probability", type="l", ylim=c(0.3, 1), lwd=2, col=24)
lines(coarsenedData$breaks, c(1, ipw_sim$S0), col=367)
lines(coarsenedData$breaks, c(1, aipw_sim$S0), col=562)
lines(coarsenedData$breaks, c(1, tmle_sim$S0), col=413)
lines(coarsenedData$breaks, c(1, cox_sim$S0), col=547)
lines(coarsenedData$breaks, c(1, weightedCox_sim$S0), col="green4")
```

```{r}
par(mfrow=c(1,2))
plot(coarsenedData$breaks, c(0, (Truth_S$S1-Truth_S$S0)), main="Risk difference", xlab="Time (days)", ylab="Difference in survival", type = "b", pch = 20, ylim=c(-0.2, 0.05), lwd=2, col=24)
lines(coarsenedData$breaks, c(0, ipw_sim$risk_diff), col=367, type = "b", pch = 20)
lines(coarsenedData$breaks, c(0, aipw_sim$risk_diff), col=562, type = "b", pch = 20)
lines(coarsenedData$breaks, c(0, tmle_sim$risk_diff), col=413, type = "b", pch = 20)
lines(coarsenedData$breaks, c(0, cox_sim$risk_diff), col=547, type = "b", pch = 20)
lines(coarsenedData$breaks, c(0, weightedCox_sim$risk_diff), col="green4", type = "b", pch = 20)
legend("bottomleft", c("Truth","IPW", "Augmented IPW", "TMLE", "Stratified cox", "Weighted stratified cox"),col=c(24, 367, 562, 413, 547, "green4"), box.lty=0, lwd=c(2, 1, 1, 1, 1, 1), cex=0.6)
plot(coarsenedData$breaks[-1], ipw_sim$SE_risk_diff, main="SD of risk difference", xlab="Time (days)", ylab="Standard deviation", type = "p", pch = 20, ylim=c(0, 0.09), col=367)
lines(coarsenedData$breaks[-1], aipw_sim$SE_risk_diff, col=562, type = "p", pch = 20)
lines(coarsenedData$breaks[-1], tmle_sim$SE_risk_diff, col=413, type = "p", pch = 20)
lines(coarsenedData$breaks[-1], cox_sim$SE_risk_diff, col=547, type = "p", pch = 20)
lines(coarsenedData$breaks[-1], weightedCox_sim$SE_risk_diff, col="green4", type = "p", pch = 20)
```

#### Difference in RMST

```{r}
par(mfrow=c(1,2))
plot(coarsenedData$breaks, c(0, ipw_sim$rmst_diff), main="Difference in RMST", xlab="Time (days)", ylab="Difference in RMST", type = "b", pch = 20, ylim=c(-0.2, 0.1), lwd=1, col=367)
lines(coarsenedData$breaks, c(0, aipw_sim$rmst_diff), col=562, type = "b", pch = 20)
lines(coarsenedData$breaks, c(0, tmle_sim$rmst_diff), col=413, type = "b", pch = 20)
lines(coarsenedData$breaks, c(0, cox_sim$rmst_diff), col=547, type = "b", pch = 20)
lines(coarsenedData$breaks, c(0, weightedCox_sim$rmst_diff), col="green4", type = "b", pch = 20)
legend("bottomleft", c("IPW", "Augmented IPW", "TMLE", "Stratified cox", "Weighted stratified cox"),col=c(367, 562, 413, 547, "green4"), box.lty=0, lwd=c(1, 1, 1, 1, 1), cex=0.6)
plot(coarsenedData$breaks[-1], ipw_sim$SE_rmst_diff, main="SD of difference in RMST", xlab="Time (days)", ylab="Standard deviation", type = "p", pch = 20, ylim=c(0, 0.09), col=367)
lines(coarsenedData$breaks[-1], aipw_sim$SE_rmst_diff, col=562, type = "p", pch = 20)
lines(coarsenedData$breaks[-1], tmle_sim$SE_rmst_diff, col=413, type = "p", pch = 20)
lines(coarsenedData$breaks[-1], cox_sim$SE_rmst_diff, col=547, type = "p", pch = 20)
lines(coarsenedData$breaks[-1], weightedCox_sim$SE_rmst_diff, col="green4", type = "p", pch = 20)
```












